Left-linear rewriting in adhesive categories

### 2024-04-22 concur/fusione.tex
minor

A noteworthy advantage of DPO rewriting is that it can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}. 
%
Operating within the framework of adhesive categories allows one to
devise the foundations of a rewriting theory that can be subsequently
instantiated to the context of interest, thus avoiding the need of
repeatedly proving similar ad-hoc results for each specific
setting. In fact, the framework is so general that it allowed to
recast in the DPO approach many additional formalisms, ranging from
toposes~\cite{johnstone2007quasitoposes} to string
diagrams~\cite{bonchi2022string}.

@@

On the one hand, they capture most categories of
interest for rewriting. In particular, we identify a vast class of
left-linear rewriting systems on graph-like structures that are
well-switching and we show that rewriting systems on graphs with
equivalences are so.\todo{weaker statement?}
As a sanity check, we prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

@@

    We call a $\mor(\X)$-Van Kampen ($\mor(\X)$-stable) square
    % simply
    a \emph{Van Kampen} (\emph{stable}) square.
  \end{definition}

@@

    \caption{Rules of the rewriting system grammar of the example.}
    \label{fi:rules}
  \end{figure}

@@

Along the lines of \cite{baldan2011adhesivity} we can identify a large
class of adhesive categories such that all left-linear rewriting
systems over such categories are strong enforcing. The details are in
\Cref{app:fill}. Here it is important only to remark that such class
includes $\cat{Set}$ and it is closed by comma and functor category
constructions.  As such, it includes essentially all categories (e.g.,
presheaves over set) that are typically considered for modelling
purposes. Notably, it contains the category $\cat{Graph}$ of directed
graphs.\todo{P: un reviewer dice ``it is unfortunate that you did not find space for describing the class in any detail, but please at least indicate how it compare to the class studied in [6].''. Mi pare sia chiaro, ma ...}

Still, there are adhesive rewriting systems that are not strong enforcing, as shown by the following example, which is however somehow pathological (recall that we fix $\mathcal{M}=\mathsf{Iso(\X)}$).

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not strong, and thus of a
  non-strong enforcing left-linear adhesive rewriting system.\todo{P: rev2 dice "I don't think that's quite right. The category in Ex. 3.7 was not adhesive, wasn't it? It was only iso-adhesive, which is a non-condition.". Mi pare chiaro, ma se viene in mente un modo di esserlo di piu' ...}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

% The previous definition might sound restrictive.
Clearly, %It is immediate to see that 
linear rewriting systems are well-switching (see Proposition~\ref{pr:weak}).

Moreover, we next observe that various classes of rewriting systems,
comprising all the ones used in modelling the applications to the
encoding of process calculi or of bio and chemical systems mentioned
in the introduction, are actually well-switching.

@@

This result fails when considering left-linear
derivations. This is readily seen by considering derivations $\der{D}$
in Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. We
already observed that $\der{D}$ and $\der{D}'$ are not switch equivalent.
% 
Still, it is easy to see that they are related by a consistent
permutation, which is the identity on productions. The colimit of
both derivations is a single node with a self loop, i.e.~the
terminal object in $\cat{Graph}$, hence all commutation requirements
are trivially satisfied.

@@

\begin{example}
  \label{ex:non-global}
  Consider the rewriting system used as running example (rules in
  Fig.~\ref{fi:rules}) and add the rules $\rho_3$ and $\rho_4$ in
  Fig.~\ref{fi:rules-cas}.\todo{P: rev.2 dice ``In the top-left and top-middle nodes of $\rho_3$, the labels "12" should be "1" and "2".''. Non e' corretto, ma magari si capisce poco ...}
  Consider the derivation $\der{F}$ in
  Fig.~\ref{fi:derF}. Note that rule $\rho_4$ needs a self-loop on
  node $12$ to be applied. This can be generated either by $\rho_3$ or
  by $\rho_2$ merging nodes $1$ and $2$, hence at least one of them
  must precede the application of $\rho_4$. Indeed, in $\der{F}$ it
  can be easily seen that the second and the third step applying
  $\rho_2$ and $\rho_4$ are independent. However, we could switch
  twice the application of $\rho_3$, bringing it in the last position,
  obtaining a derivation $\der{F}'$ as depicted in
  Fig.~\ref{fi:derF1}, where the applications of $\rho_2$ and $\rho_4$
  are no longer independent.

@@

While inversions can be
performed in any order in the case of linear rewriting systems, this is no longer true for left-linear
rewriting systems due to the fact that, as already observed, globality of independence holds in a weak form. For instance, if we get back to Example~\ref{ex:non-global} and consider derivation $\der{F}$ in Fig.~\ref{fi:derF}, the last two steps applying rules $\rho_2$ and $\rho_4$ can be switched, thus leading to a derivation, call it $\der{F}''$,  applying $\rho_3, \rho_4, \rho_2$. Such $\der{F}$ is switch equivalent to $\der{F}'$ in Fig.~\ref{fi:derF1}, hence starting from $\der{F}''$ we can obtain $\der{F}'$ via a switching sequence, but we cannot start by switching $\rho_2$ and $\rho_4$, despite the fact that this is an inversion.

@@

We performed an in-depth investigation of the notion of independence between
rewriting steps in the setting of left-linear rewriting systems
over adhesive categories, which encompasses most of the
structures commonly used in analysing and modelling applications.

@@

The class of well-switching adhesive rewriting systems is large. Generalising a result in~\cite{baldan2011adhesivity},
we showed that one of its defining properties (the fact that sequential independence implies switchability) holds 
for a general class of adhesive categories that include all presheaves over $\cat{Set}$. Moreover, the second property, uniqueness 
of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the sorts 
of depth 0  are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it 
happens in most of the approaches to the modelling of calculi and biological systems that we cited in the introduction, 
as well as in string diagrams).
%\todo{Giusto, refernences?}

The results in this paper open the way to the development of a
concurrent semantics for left-linear rewriting systems over adhesive
categories. In~\cite{baldan2017domains} the authors argue that for
computational systems that allow one to express merging or fusions,
the right semantical models are weak prime domains, a generalisation
of prime domains, a staple in concurrency theory, and connected event
structures, their event-based counterpart. The mentioned paper
discusses only the case of graph rewriting and does not focus into the
problems induced by the occurring of ``merges'' to the theory of rewriting (the systems
considered there are implicitly assumed to be well-switching). 
We plan to consolidate the statement that weak prime domains are 
``the'' model for systems with fusions by showing that they allow to provide a semantics
for left-linear rewriting systems in $\mathcal{M}$-adhesive categories.

>>>>

A noteworthy advantage of DPO rewriting is that it can be formulated over
adhesive categories and their variants~\cite{lack2005adhesive,ehrig2006weak}, which include, among others, 
toposes~\cite{johnstone2007quasitoposes} and string
diagrams~\cite{bonchi2022string}.
%
Operating within the framework of adhesive categories allows one to
devise the foundations of a rewriting theory that can be then
instantiated to the context of interest, thus avoiding the need of
repeatedly proving similar ad-hoc results for each specific
setting. % In fact, the framework is so general that it allowed to
% recast in the DPO approach many additional formalisms, ranging from
% toposes~\cite{johnstone2007quasitoposes} to string
% diagrams~\cite{bonchi2022string}.

@@

On the one hand, they capture most categories of
interest for rewriting. Indeed, we identify quite general classes of
left-linear rewriting systems on graph-like structures that are
well-switching.
% and we show that rewriting systems on graphs with
% equivalences are so.
As a sanity check, we also prove that linear rewriting systems on
$\mathcal{M}$-adhesive categories are well-switching.

@@

    A $\mor(\X)$-Van Kampen square will be called simply \emph{Van
      Kampen} and similarly, a $\mor(\X)$-stable square is called
    \emph{stable}.
  \end{definition}

@@

    \caption{A rewriting system $\cat{Graph}$.}
    \label{fi:rules}
  \end{figure}

@@

We can identify a large class of adhesive categories such that all
left-linear rewriting systems over such categories are strong
enforcing. Such class includes $\cat{Set}$ and it is closed
by comma and functor category constructions.  As such, it includes
essentially all categories (e.g., presheaves over set) that are
typically considered for modelling purposes. Notably, it contains the
category $\cat{Graph}$ of directed graphs.
This is a natural generalisation from adhesive to $\mathcal{M}$-adhesive catgeories of a class studied in \cite{baldan2011adhesivity} (see \Cref{app:fill} for details).

Still, there are $\mathcal{M}$-adhesive rewriting systems that are not strong enforcing, as shown by the following example.

\begin{example}[Non-strong enforcing left-linear rewriting system]
  \label{ex:diff2}
  In light of \Cref{pr:church}, \Cref{ex:diff1} provides an example of
  an independence pair which is not strong. This gives an example of a
  left-linear rewriting system in a $\mathcal{M}$-adhesive category
  which is not strong enforcing. The example is somehow
  pathological since $\mathcal{M}=\mathsf{Iso(\X)}$, but it needs to be so, since natural examples are all in the well-beahved class mentioned above. 
  .\todo{P: Scritto qualcosa, probabilmente da migliorare}
  % Instead every
  % left-linear system in category $\cat{Graph}$ is strong enforcing, by
  % the results of Appendix~\ref{app:fill}, summarized
  % in~\Cref{ex:graph-type-bb} and~\Cref{cor:llr-is-strEnf}.
\end{example}

@@

% The previous definition might sound restrictive.
Clearly, %It is immediate to see that 
linear rewriting systems are well-switching (see Proposition~\ref{pr:weak}).
%
Moreover, we next observe that various classes of rewriting systems,
comprising all the ones used in modelling the applications to the
encoding of process calculi or of bio and chemical systems mentioned
in the introduction, are actually well-switching.

@@

This result fails when considering left-linear
derivations. This is readily seen by considering derivations $\der{D}$
in Fig.~\ref{fi:derD} and $\der{D}'$ in Fig.~\ref{fi:derD1}. We
observed that $\der{D}$ and $\der{D}'$ are not switch equivalent.
% 
Still, it is easy to see that they are related by a consistent
permutation, which is the identity on productions. The colimit of
both derivations is a single node with a self loop, i.e.~the
terminal object in $\cat{Graph}$, hence all commutativity requirements
are trivially satisfied.

@@

\begin{example}
  \label{ex:non-global}
  Consider a new rewriting system in $\cat{Graph}$ consisting of the
  rules in $\rho_0$, $\rho_1$ and $\rho_2$ in
  Fig.~\ref{fi:rules-cas}.\todo{P: rev.2 dice ``In the top-left and
    top-middle nodes of $\rho_3$, the labels "12" should be "1" and
    "2".''. Non e' corretto, ma magari si capisce poco ...}  Consider
  the derivation $\der{F}$ in Fig.~\ref{fi:derF}. Note that rule
  $\rho_2$ needs the black self-loop on node $12$ to be applied. This can be
  generated either by $\rho_0$ or by $\rho_1$ merging nodes $1$ and
  $2$, hence at least one of them must precede the application of
  $\rho_2$. Indeed, in $\der{F}$ it can be easily seen that the second
  and the third step applying $\rho_1$ and $\rho_2$ are
  independent. However, we could switch twice the application of
  $\rho_0$, bringing it in the last position, obtaining a derivation
  $\der{F}'$ as depicted in Fig.~\ref{fi:derF1}, where the
  applications of $\rho_1$ and $\rho_2$ are no longer independent.

@@

While inversions can be
performed in any order in the case of linear rewriting systems, this is no longer true for left-linear
rewriting systems due to the fact that, as already observed, globality of independence holds in a weak form. For instance, if we get back to Example~\ref{ex:non-global} and consider derivation $\der{F}$ in Fig.~\ref{fi:derF}, the last two steps applying rules $\rho_1$ and $\rho_2$ can be switched, thus leading to a derivation, call it $\der{F}''$,  applying $\rho_0, \rho_2, \rho_1$. Such $\der{F}''$ is switch equivalent to $\der{F}'$ in Fig.~\ref{fi:derF1}, hence starting from $\der{F}''$ we can obtain $\der{F}'$ via a switching sequence, but we cannot start by switching $\rho_0$ and $\rho_2$, despite the fact that this is an inversion.

@@

We performed an in-depth investigation of the notion of independence between
rewriting steps in the setting of left-linear rewriting systems
over $\mathcal{M}$-adhesive categories, which encompasses most of the
structures commonly used in analysing and modelling applications.

@@

The class of well-switching $\mathcal{M}$-adhesive rewriting systems is large. Generalising a result in~\cite{baldan2011adhesivity},
we showed that one of its defining properties (the fact that sequential independence implies switchability) holds 
for a general class of $\mathcal{M}$-adhesive categories that include all presheaves over $\cat{Set}$. Moreover, the second property, uniqueness 
of switching, is ensured when rewriting possibly hierarchical graphical structures, as long as elements of the sorts 
of depth 0  are not merged (e.g., one can consider rewriting systems over directed graphs where only nodes are merged, as it 
happens in most of the approaches to the modelling of calculi and biological systems that we cited in the introduction, 
as well as in string diagrams).
%\todo{Giusto, refernences?}

The results in this paper open the way to the development of a
concurrent semantics for left-linear rewriting systems over $\mathcal{M}$-adhesive
categories. In~\cite{baldan2017domains} the authors argue that for
computational systems that allow one to express merging or fusions,
the right semantical models are weak prime domains, a generalisation
of prime domains, a staple in concurrency theory, and connected event
structures, their event-based counterpart. The mentioned paper
discusses only the case of graph rewriting and does not focus into the
problems induced by the occurring of ``merges'' to the theory of rewriting (the systems
considered there are implicitly assumed to be well-switching). 
We plan to consolidate the statement that weak prime domains are 
``the'' model for systems with fusions by showing that they allow to provide a semantics
for left-linear rewriting systems in $\mathcal{M}$-adhesive categories.
